Sleep, relative to wake, increases both veridical and false memory in the DRM paradigm: A registered report

### 2023-07-10 manuscript.tex
Minor edits and fixes

\usepackage{caption}

@@

In this section, we present the results of four exploratory analyses, which explored {[}1{]} the number of total responses (i.e., studied + lure + intrusions) across groups, {[}2{]} whether the effect of sleep on lure recall is modulated by veridical recall, as suggested by Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]}, {[}3{]} the extent to which a lure being produced is predicted by its corresponding list items being recalled, and {[}4{]} the semantic distance between intrusions and critical lures.

@@

In this exploratory analysis, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample's adjusted veridical recall (Median = 5), classifying our participants as either high (\textgreater5; N = 117) or low (\(\leq\) 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]}, but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs.~Wake) \(\times\) 2 (Adjusted veridical recall: High vs.~Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group (\(\beta\) = 0.052, z = 0.494, p = .621) but a main effect of Adjusted veridical recall (\(\beta\) = -0.438, z = -3.766, p \textless{} .001), such that high performers (M = 3.17, SD = 2.71) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.68) ({[}\protect\hyperlink{ref-thapar2001a}{40},\protect\hyperlink{ref-toglia1999a}{41}{]}, but see {[}\protect\hyperlink{ref-roediger2001a}{24},\protect\hyperlink{ref-cann2011a}{42},\protect\hyperlink{ref-stadler1999a}{43}{]}). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant (\(\beta\) = 0.049, z = 0.30, p = .763). Despite this, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (M\(_{Sleep}\) = 2.09, SD\(_{Sleep}\) = 1.63 vs.~M\(_{Wake}\) = 2.09, SD\(_{Wake}\) = 1.74; \(\beta\) = 0.133, z = 0.021, p = .982). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.'s {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} findings.

@@

As per the DRM literature, responses that were neither the studied list words nor the critical lures were classified as (non-critical) intrusions. For instance, our participants studied nurse, sick, lawyer, medicine, health, hospital, dentist, and physician, with doctor as the critical lure. Responses such as clinic and coconut would both be considered intrusions, but clearly, clinic is semantically more related to the studied list words than coconut. In other words, there is much diversity within the intrusion data. Here, we explored whether our four groups differed in terms of the semantic distance between intrusions and critical lures, as indexed by pre-trained semantic spaces (ukWaC; {[}\protect\hyperlink{ref-baroni2009a}{45}{]}) derived from word2vec {[}\protect\hyperlink{ref-guenther2015a}{46}{]}. We reasoned that since lure recall was greater in our Sleep (vs.~Wake) participants, the intrusions produced by these participants could be more related to the lures in semantic space (e.g., {[}\protect\hyperlink{ref-mak2023a}{15}{]}). To test this, we computed the cosine similarities between each intrusion and each of the 20 critical lures (see Table \ref{tab:table7} for an illustration). The intrusion-lure pair with the highest cosine similarity (i.e., the nearest neighbour) was used for this analysis.

@@

In light of the extensive dataset and the comprehensive analyses conducted, we decided to summarise our key findings in Table 8 to help readers gain a better understanding of the overall picture.

@@

@@

@@

`According to Newbury and Monaghan's {[}\protect\hyperlink{ref-newbury2019a}{1}{]} meta-analysis, the effect size for sleep in DRM false memory is Hedge's g = +0.92 (95\% CI: 0.54, 1.30; p \textless{} .001) in short lists (10 words per list).\footnote{Hedge’s g and Cohen’s d are interchangeable when the sample size is larger than 30 (Kline, 2004; Lakens, 2013).} However, due to certain biases in the psychology literature (e.g., publication bias), it has been argued that published effect sizes are generally inflated {[}e.g., \protect\hyperlink{ref-schaefer2019a}{59}{]} and that power analysis should be based on the lowest meaningful estimate {[}\protect\hyperlink{ref-albers2018a}{60},\protect\hyperlink{ref-cortex2013a}{61}{]}. Therefore, we opted for a more conservative (yet contextualised) effect size estimate and went for the lower-bound of the 95\% confidence interval reported by Newbury and Monaghan {[}\protect\hyperlink{ref-newbury2019a}{1}{]}, which is g = +0.54.

@@

We then merged the two fabricated datasets together and fitted a GLMM to it, using the lme4 package {[}\protect\hyperlink{ref-bates2015a}{63}{]}. The dependent variable, fixed effects structure, coding scheme, and computation procedures were identical to those described in section 12.3. Table D1 shows the fixed-effects estimates from the converged model, which has a by-participant intercept only.

Table D1

Based on the fixed-effects estimates, we estimated the power a sample size of 480 (i.e., 120/group) has for detecting an Interval and Test Time interaction. We conducted Monte Carlo simulation using the ``simr'' package {[}\protect\hyperlink{ref-green2016a}{64}{]} in R (see Box 2 for R codes). After 500 simulations, it was estimated that a sample of this size gives 90\% power (95\% CI: 87.03, 92.49) to detect a significant interaction. Then, we estimated the power we have for detecting a simple effect of Test Time within the Delay groups (i.e., Sleep vs.~Wake). Following the simulation procedures above, it was estimated that 120 participants/group (i.e., 240 in the Delay groups) will give about 90\% power (95\% CI: 97.41, 99.56). In sum, our power calculation showed that having 120 participants/groups will give ample power (\textgreater90\%) to detect both an Interval x Test Time interaction and a simple effect of Sleep vs.~Wake. Finally, we also estimated that we will have at least 80\% power as long as we have \textgreater99 participants/group. Therefore, in case we fail to reach our target sample size of 120 participants/group before funding expires but manage to recruit \textgreater99/group, our proposed experiment will still have satisfactory power.

>>>>

\usepackage{caption} \usepackage{float} \floatplacement{figure}{H} \newcommand{\beginsupplement}{\setcounter{table}{0} \renewcommand{\thetable}{D\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{D\arabic{figure}}}

@@

In this section, we present the results of four exploratory analyses, which explored (1) the number of total responses (i.e., studied + lure + intrusions) across groups, (2) whether the effect of sleep on lure recall is modulated by veridical recall, as suggested by Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]}, (3) the extent to which a lure being produced is predicted by its corresponding list items being recalled, and (4) the semantic distance between intrusions and critical lures.

@@

In this exploratory analysis, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample's adjusted veridical recall (Median = 5), classifying our participants as either high (\textgreater5; N = 117) or low (\(\leq\) 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]}, but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs.~Wake) \(\times\) 2 (Adjusted veridical recall: High vs.~Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group (\(\beta\) = 0.052, z = 0.494, p = .621) but a main effect of Adjusted veridical recall (\(\beta\) = -0.438, z = -3.766, p \textless{} .001), such that high performers (M = 3.24, SD = 2.63) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.63) ({[}\protect\hyperlink{ref-thapar2001a}{40},\protect\hyperlink{ref-toglia1999a}{41}{]}, but see {[}\protect\hyperlink{ref-roediger2001a}{24},\protect\hyperlink{ref-cann2011a}{42},\protect\hyperlink{ref-stadler1999a}{43}{]}). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant (\(\beta\) = -0.025, z = -0.3, p = \textless.0001). Despite this, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (M\(_{Sleep}\) = 2.09, SD\(_{Sleep}\) = 1.63 vs.~M\(_{Wake}\) = 2.09, SD\(_{Wake}\) = 1.74; \(\beta\) = 0.003, z = 0.021, p = .983). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.'s {[}\protect\hyperlink{ref-diekelmann2010a}{39}{]} findings.

@@

As per the DRM literature, responses that were neither the studied list words nor the critical lures were classified as (non-critical) intrusions. For instance, our participants studied nurse, sick, lawyer, medicine, health, hospital, dentist, and physician, with doctor as the critical lure. Responses such as clinic and coconut would both be considered intrusions, but clearly, clinic is semantically more related to the studied list words than coconut. In other words, there is much diversity within the intrusion data. Here, we explored whether our four groups differed in terms of the semantic distance between intrusions and critical lures, as indexed by pre-trained semantic spaces (ukWaC; {[}\protect\hyperlink{ref-baroni2009a}{45}{]}) derived from word2vec {[}\protect\hyperlink{ref-guenther2015a}{46}{]}. We reasoned that since lure recall was greater in our Sleep (vs.~Wake) participants, the intrusions produced by these participants could be more related to the lures in semantic space (e.g., {[}\protect\hyperlink{ref-mak2023a}{15}{]}). To test this, we computed the cosine similarities between each intrusion and each of the 20 critical lures (see Table \ref{tab:table8} for an illustration). The intrusion-lure pair with the highest cosine similarity (i.e., the nearest neighbour) was used for this analysis.

\caption{\label{tab:table8}Procedure for the exploratory analysis on semantic distance between intrusion and lure.}
\centering
\begin{tabular}[t]{llll}
\toprule
\textbf{Participant ID} & \textbf{Intrusion produced} & \textbf{Lure} & \textbf{cosine}\\
\textbf{} & \textbf{by a participant} & \textbf{} & \textbf{}\\
\midrule
1 & ear & smell & 0.405\\
1 & ear & sleep & 0.302\\
1 & ear & doctor & 0.239\\
\addlinespace
1 & ear & window & 0.129\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering\begingroup\fontsize{8}{10}\selectfont

\begin{tabular}{llllcc}
\toprule
\textbf{Participant ID} & \textbf{Intrusion produced} & \textbf{Closest lure} & \textbf{cosine} & \textbf{Number of intrusion} & \textbf{Average cosine/}\\
\textbf{} & \textbf{by a participant} & \textbf{} & \textbf{} & \textbf{produced} & \textbf{participant}\\
\midrule
1 & ear & smell & 0.405 &  & \\
1 & metro & city & 0.351 & 4 & 0.347\\
1 & heavy & slow & 0.413 &  & \\
\addlinespace
1 & clear & slow & 0.219 &  & \\
\midrule
2 & clinic & doctor & 0.494 &  & \\
2 & beach & mountain & 0.484 & 2 & 0.489\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

In light of the extensive dataset and the comprehensive analyses conducted, we decided to summarise our key findings in Table \ref{tab:table9} to help readers gain a better understanding of the overall picture.

\caption{\label{tab:table9}Summary of the key findings.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llllll}
\toprule
\textbf{Categories} & \textbf{Interval x} & \textbf{Mean in the} & \textbf{Mean in the} & \textbf{Pairwise} & \textbf{Interpretation}\\
\textbf{} & \textbf{Test Time} & \textbf{Sleep group} & \textbf{Wake group} & \textbf{comparison} & \textbf{}\\
\textbf{} & \textbf{p<.05?*} & \textbf{(SD)} & \textbf{(SD)} & \textbf{p<.05?} & \textbf{}\\
\midrule
Intrusions & Yes & 6.78 (8.71) & 10.1 (13.65) & Yes & Few intrusions\\
 &  &  &  &  & after sleep\\
\midrule
\addlinespace
Critical lures & Yes (N of intrusions & 2.72 (2.3) & 2.51 (2.32) & Yes & Greater DRM false\\
(Max=20) & included as a coviariate) &  &  &  & recall after sleep\\
\midrule
Studied list words & Yes (N of intrusions & 17.32 (16.59) & 13.41 (12.93) & Yes & Greater veridical recall\\
(Max=160) & included as a coviariate) &  &  &  & after sleep\\
\midrule
Total responses & No & 26.82 (18.63) & 26.02 (21.49) & No & Sleep and Wake groups\\
\addlinespace
 &  &  &  &  & were well-matched\\
\bottomrule
\multicolumn{6}{l}{\textsuperscript{} * A significant Interval x Test Time interaction suggests that any difference between the Sleep and Wake groups cannot}\\
\multicolumn{6}{l}{\textsuperscript{} be solely explained by study/test being completed at different time of day.}\\
\end{tabular}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{l}
\toprule
\multicolumn{1}{c}{Page 1: Demographic information}\\
\midrule
1)  What is your gender identity?\\
2)  How old are you?\\
3)  In what country do you currently live?\\
4)  What is your first language(s)?\\
\addlinespace
5)  What is your ethnicity?\\
6)  What is the highest level of education you have completed?\\
7)  Do you have any history of any psychiatric (e.g., schizophrenia), developmental\\
(e.g., autism, dyslexia), or sleep disorders (e.g., insomnia)?\\
8)  If your answer to the above is Yes, please can you name the diagnosis/es?\\
\midrule
\addlinespace
\multicolumn{1}{c}{Page 2: Outline of the main study}\\
\midrule
\textbf{IMPORTANT: Please read carefully}\\
We are recruiting hundreds of participants for a simple memory study. We would\\
like to see if you may be interested in taking part.\\
In Task 1, participants will see and remember some English words. This will take about 10 mins.\\
\addlinespace
In Task 2, participants will complete a simple memory test based on the words they saw in Task 1.\\
This requires about 12 mins.\\
Participants will receive £3.5 (£9.5/hour) upon completion of the two tasks.\\
Importantly, participants will be randomly allocated to one of the four groups:\\
Group A (AM Group): You can start Task 1 and 2 any time between 8.30-10.30AM.\\
\addlinespace
Group B (PM Group): You can start Task 1 and 2 any time between 8.30-10.30PM.\\
Group C (Delay Group 1): You can start Task 1 any time between 8.30-10.30AM and then Task 2\\
between 8.30-10.30PM on the same day.\\
Group D (Delay Group 2): You can start Session 1 any time between 8.30-10.30PM and then Task 2\\
between 8.30-10.30AM the day after.\\
\addlinespace
Those in Groups C and D will receive a £0.2 bonus upon completion of the study. Unfortunately,\\
we are NOT able to accommodate any preferences for group allocation.\\
If you are happy to take part in our memory study, press Yes below. If not, press No.\\
Yes / No\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  (Sleep group only) Approximately what time did you go to bed last night?
\item
  (Sleep group only) Approximately what time did you wake up this morning?
\item
  (Sleep group only) How would you rate the quality of last night's sleep?
  \textbf{Very Good, Good, Fair, Poor, Very Poor}
\item
  (Wake group only) Did you have a nap between Session 1 and now?
  \textbf{Yes, No}
\item
  (Wake group only) If your answer to the above is Yes, how long was the nap?
\item
  Did you consume any alcoholic drinks in the last 12 hours? If Yes, how much?
  Yes, No
\item
  Did you consume any caffeinated drink in the last 6 hours? If Yes, how much?
  Yes, No
\item
  How many people are in close proximity (\textless{} 3 meter) to you RIGHT NOW?
\item
  How bright is your immediate environment RIGHT NOW?
  \textbf{Too bright, Sufficiently bright, A bit dark, Very dark}
\item
  How noisy is the environment you are in RIGHT NOW?
  \textbf{Very quiet, Quiet, Noisy, Very noisy.}
\end{enumerate}

\textbf{The following questions were taken from the reduced version of the Morningness/Evenningess questionnaire (Adan \& Almirrall, 1991).}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{10}
\item
  Approximately what time would you get up if you were entirely free to plan your day?
  \textbf{5am-6:30am, 6:30am-7:45am, 7:45am-9:45am, 9:45am-11am, 11am-12 noon}
\item
  On a regular day, during the first half hour after you wake up in the morning, how do you feel?
  \textbf{Very tired, Fairly tired, Fairly refreshed, Very refreshed}
\item
  On a regular day, at approximately what time in the evening do you feel tired, and, as a result, in need of sleep?
  \textbf{8-9pm, 9-10:15pm, 10:15pm-12:45am, 12:45-2am, 2-3am}
\item
  On a regular day, at approximately what time of day do you usually feel your best?
  \textbf{5-8am, 8-10am, 10am-5pm, 5-10pm, 10pm-5am}
\item
  One hears about ``morning types'' and ``evening types.'' Which one of these types do you consider yourself to be?
  \textbf{Definitely an evening type, Rather more an evening type than a morning type, Rather more a morning type than an evening type, Definitely a morning type}
\end{enumerate}

@@

\setcounter{table}{0} \renewcommand{\thetable}{D\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{D\arabic{figure}}

According to Newbury and Monaghan's {[}\protect\hyperlink{ref-newbury2019a}{1}{]} meta-analysis, the effect size for sleep in DRM false memory is Hedge's g = +0.92 (95\% CI: 0.54, 1.30; p \textless{} .001) in short lists (10 words per list).\footnote{Hedge’s g and Cohen’s d are interchangeable when the sample size is larger than 30 (Kline, 2004; Lakens, 2013).} However, due to certain biases in the psychology literature (e.g., publication bias), it has been argued that published effect sizes are generally inflated {[}e.g., \protect\hyperlink{ref-schaefer2019a}{59}{]} and that power analysis should be based on the lowest meaningful estimate {[}\protect\hyperlink{ref-albers2018a}{60},\protect\hyperlink{ref-cortex2013a}{61}{]}. Therefore, we opted for a more conservative (yet contextualised) effect size estimate and went for the lower-bound of the 95\% confidence interval reported by Newbury and Monaghan {[}\protect\hyperlink{ref-newbury2019a}{1}{]}, which is g = +0.54.

@@

We then merged the two fabricated datasets together and fitted a GLMM to it, using the lme4 package {[}\protect\hyperlink{ref-bates2015a}{63}{]}. The dependent variable, fixed effects structure, coding scheme, and computation procedures were identical to those described in section 12.3. Table \ref{tab:tableD1} shows the fixed-effects estimates from the converged model, which has a by-participant intercept only.

\caption{\label{tab:tableD1}Fixed-effects estimates from the converged maximal model examining the effects of Interval and Test Time in the fabricated dataset.}
\centering
\begin{tabular}[t]{lllll}
\toprule
\textbf{Fixed effects} & \textbf{Estimate} & \textbf{SE} & \textbf{z} & \textbf{p}\\
\midrule
Intercept & -0.316 & 0.024 & -13.244 & <.001\\
Interval & -0.039 & 0.024 & -1.65 & .099\\
Test Time & 0.063 & 0.024 & 2.663 & .008\\
Interval x Test Time & 0.076 & 0.024 & 3.198 & .001\\
\bottomrule
\end{tabular}
\end{table}

Based on the fixed-effects estimates, we estimated the power a sample size of 480 (i.e., 120/group) has for detecting an Interval and Test Time interaction. We conducted Monte Carlo simulation using the ``simr'' package {[}\protect\hyperlink{ref-green2016a}{64}{]} in R (see Box 2 for R codes). After 500 simulations, it was estimated that a sample of this size gives 90\% power (95\% CI: 87.03, 92.49) to detect a significant interaction. Then, we estimated the power we have for detecting a simple effect of Test Time within the Delay groups (i.e., Sleep vs.~Wake). Following the simulation procedures above, it was estimated that 120 participants/group (i.e., 240 in the Delay groups) will give about 57.2\% power (95\% CI: 52.73, 61.58). In sum, our power calculation showed that having 120 participants/groups will give ample power (\textgreater90\%) to detect both an Interval x Test Time interaction and a simple effect of Sleep vs.~Wake. Finally, we also estimated that we will have at least 80\% power as long as we have \textgreater99 participants/group. Therefore, in case we fail to reach our target sample size of 120 participants/group before funding expires but manage to recruit \textgreater99/group, our proposed experiment will still have satisfactory power.
